\LARGE \bf Machine Translation for Tibetan Buddhist Texts

### 2021-04-16 paper/main.tex
Update main.tex
Fixed citation typos, removed appendix sentence, and other small changes

\thanks{$^{3}$ Rahul Sehrawat Columbia '21
        {\tt\small rs3688 at columbia dot edu}}%

@@

@@

We have trained our models in PyTorch and have leveraged the BART architecture in the HuggingFrance transformers library to make the training and architecture more straightforward. By creating these models, we hope to motivate scholars to study not only these texts but Tibetan as a whole. Having a readily available set of translated discourses could create an on-ramp for further language uses because increased access could lead to increased interest, especially in the English-speaking community. 
Currently, while there has been some interest in machine translation for Tibetan-Chinese and vice versa, the same interest has not been reflected in Tibetan-English machine translation. We attempted to find open-source Tibetan NLP material, but we were unable to find any projects attempting to translate Tibetan-English nor could we find papers specifically discussing Tibetan-English machine translation. While our model tackles this difficult task in creating a translator for Tibetan-English, it is by no means a perfect or even accurate model. In the final iteration of our model, we were only able to achieve a 23.01\% BLEU score, which means that our model has low-quality output.

This is caused by several factors, from the difficulty to preprocess Tibetan sentences to the lack of Tibetan-English data freely available online and our overall knowledge of the Tibetan language. While it may seem that we were unable to fully accomplish our lofty goal of creating a Tibetan-English translator it’s important to note that we are treading in nearly uncharted territory. We hope that our endeavor may help future projects in this field and provide insight into the current state of Tibetan-English machine translation.

\subsection{RELATED WORK}

@@

To train our model, we acquired publicly available English to Tibetan parallel corpora. The non-profit 84000 provided 80 English to Tibetan parallel texts on their website \cite{c8}. All 80 of these texts were expert translations of Indo-Tibetan Buddhist texts. For our model we used all 80 of these provided texts.

Before using 84000's corpora however, we pre-processed the provided data to meet our model's need. The provided files from 84000 were in a .tmx file format, a very standard parallel corpus file format \cite{c9}. 
For our approach, we scraped all 80 .tmx files for both its English and Tibetan contents and placed them line by line into two large .txt files. One .txt file contained the English lines, and the other .txt file contained the Tibetan translation. We made careful assurance that the lines remained in exactly one to one correspondence with their translations. That is, a newline character was used to distinguish two separate lines of translation, but the same line number in each file corresponds to the exact translation provided by 84000. Therefore, the total line numbers in the two .txt files were necessarily the same since one line in one file corresponding to exactly the same line of translated text in the other file. Two such .txt file were made for each of the 80 .tmx files, a Tibetan and an English .txt file for each. After all 160 .txt files were generated, two single large .txt file was generated by concatenating all the Tibetan .txt files and all the English .txt files. These last two .txt files are what was fed into the model.

@@

\subsection{Model Design}
We used a 6 layer Transformer model that used a multi headed attention mechanism and a switchout transform. Each layer of the transformer model can be divided into an encoder and decoder layer. The encoding layer consists first of a self attention layer, an attention aspect which focuses on the encoded sentence, and then outputs to a feed forward layer. The multi part of the multi headed attention means that each head will encode the information in parallel and then the results will be concatenated and passed to the decoder. Figure one demonstrates this process.
\includegraphics[scale=.4]{Transformer.png}
\caption{\\Figure 1: Overview of a Transformer model \cite{11}}
\\
\par
To prevent overfitting and to augment the data, we added in a switchout transform which will randomly switch out words in the source and target sentence. This also has the added advantage of augmenting the data even further which is especially important for our problem as religious Tibetan has a relatively small corpus, around 30 thousand words. We also attempted to use masking, however we found it to be ineffective. Masking hides tokens from the model by replacing the tokens with a <mask> token. Tibetan tokens are not separated in the same way that latin languages separate their tokens. For example,
a line in our test corpus translated to “Having the altruistic intention of excellent bodhicitta.” However, the Tibetan word is only one token. Therefore, the mask would not help the model learn at all, as the entire sentence would be marked as <mask> and there would not be enough context for our model to discover the meaning. As a result, we did not use any form of masking in our model. Switchout does not create the same problem as their is always a token to take its place. \par
After implementing our transformer model with switchout we tuned hyper parameters to maximize our Transformer model. We manually tuned the following hyperparameters: dropout, label smoothing, rnn size, the number of transformer layers and the number of heads of the attention model. The final configuration used can be found in the appendix. \par
We used four different metrics to evaluate our results: accuracy, predicted score, perplexity and bilingual evaluation understudy (BLEU). The validation accuracy tells us how well we are learning our training data, the predicted score tells us the average log likelihood for each word, the perplexity tells us how well the model predicts the data and the bleu score evaluates the quality of the translation. Together, these metrics provide us with a decent picture of how well our model is performed.

@@

\par
Our six layer transformer model with switch out was unsuccessful at translating religious tibetan to english. The translated output does not match closely enough in either pure accuracy or in the meaning of the passage to be useful in translating religious tibetan texts to english. \par
While superficially, the perplexity appears to demonstrate a strong model, further investigation demonstrates that the low perplexity is a sign of inaccuracy. Perplexity measures the number of possible words that a model could predict for any given sentence. A higher perplexity indicates higher uncertainty. However, the low BLEU score, which measures the quality of the machine translation, has already demonstrated that the accuracy is low. This combined with the perplexity shows the models overconfidence in predicting certain phrases. For example, one Tibetan phrases expert translation was “So the bodhisattva mahāsattva Maitreya asked the Blessed One”, but the top five of the predicted translations all are of the form “The Blessed One then said to the [adj] [proper noun]”. Even though this sentence is wrong, the perplexity is low because the model is confident in the first part of the sentence. This phenomenon is repeated throughout the results. Keyphrases from the Tibetan sentence are interpreted wrongly in the same way, and then the model learns to be confident about its predictions which lowers the perplexity. \par
The results demonstrate the flaw in the Transformer model using the current parallel Tibetan to English corpuses. Each token in the Tibetan sentence corresponds to an entire phrase in english rather than a separate word. This complicates the translation, and the model learns parts of the phrases quite well. The validation accuracy is relatively high because the transformer model is good at learning parts of these phrases but it is not good at learning entire sentences or stringing together sentences in context which contributes to the low BLEU score. Overall, the model is capable of learning specific phrases well, but fails at the broad task of creating coherent translations from religious tibetan texts to english.

@@

Considering our use of word segmentation for preprocessing our data (as opposed to character segmentation), our results are comparable with, albeit lower than, Tibetan-Chinese models. Lai, Zhao, and Bao \cite{2} ran several different Tibetan-Chinese models on the same data using two different preprocessing methodologies, character segmentation and word segmentation. Their word segmented Transformer model averaged a higher BLEU score than ours, scoring 38.44\% while our model scored 23.01\%. While this comparison may lead to the conclusion that our model is flawed and irredeemable, only the former is correct. There’s no doubt that our model is underperforming in comparison to other transformer models and hardly makes a correct prediction at all. However, this is most likely due to our difference in preprocessing techniques and the linguistic differences and similarities between English, Chinese, and Tibetan.

Unlike English but similar to Chinese, Tibetan words are not clearly separated by spaces, rather, characters that represent words are strung together to make a sentence \cite{5}. This makes preprocessing the data and mapping Tibetan words to their English counterparts a difficult task and it makes preprocessing on characters impossible. We trained our model on a parallel corpus of English and Tibetan sentences instead of segmenting each word since preprocessing these sentences to word-to-word mappings would require extensive knowledge of the Tibetan language which our team did not have. On top of this, a Chinese-Tibetan model might perform overall better than a Tibetan-English one since Chinese and Tibetan are part of the same language family, and it is likely that characters may map to one another and the same preprocessing techniques can be used in both languages. Therefore, while comparing our model to Lai, Zhao, and Bao’s \cite{2} model may be helpful to give a general baseline on how other Tibetan translators are performing, it should not be considered where our model must perform.

Another issue with our data that may be remedied in future iterations is the corpora we chose to train and test our model on. Since we trained on sentences in our testing set, it is very possible that the model may have read some of the initial characters and assumed that the entire sentence would be a sentence from our training data that began with the same characters. This may explain why our model overpredicts the sentence “Our blessed one” with such a low perplexity. As stated above, this issue might be remedied if we could preprocess our data so that words mapped to words but since we are limited by our ability to do such, we could consider training our model on a different corpus altogether. Instead of using the translated Buddhist sutras to train our model, we could use a more general Tibetan-English corpus to train our model and then use transfer learning to fine-tune our model to religious texts. However, at this current time, the translated sutras provided by 84000 \cite{8} are some of the only parallel translations available on the internet.

Perhaps the best way to translate Tibetan to English would be to take another approach to our model design altogether and abandon the tested Transformer model. We could build an Encoder-Decoder structure based on Back Translation which would help deal with the lack of Tibetan-English translations. In this model, we would mix Tibetan and English sentences together to train the model \cite{10}. This would help to increase the size of our training data since it would no longer just be Tibetan sentences. Since this type of model has yet to be tested for Tibetan-English, it's uncertain whether this change could actually improve our translation accuracy and BLEU score, however based on papers written on Back Translation, this design looks promising.

@@

(Chris)
\begin{thebibliography}{99}

>>>>

\thanks{$^{4}$ Rahul Sehrawat Columbia '21
        {\tt\small rs3688 at columbia dot edu}}%

@@

\subsection{Overview}

@@

We have trained our models in PyTorch and have leveraged the BART architecture in the HuggingFrance transformers library to make the training and architecture more straightforward. By creating these models, we hope to motivate scholars to study not only these texts but Tibetan as a whole. Having a readily available set of translated discourses could create an on-ramp for further language uses because increased access could lead to increased interest, especially in the English-speaking community. 
Currently, while there has been some interest in machine translation for Tibetan-Chinese and vice versa, the same interest has not been reflected in Tibetan-English machine translation. We attempted to find open-source Tibetan NLP material, but we were unable to find any projects attempting to translate Tibetan-English nor could we find papers specifically discussing Tibetan-English machine translation.

While our model tackles this difficult task in creating a translator for Tibetan-English, it is by no means a perfect or even accurate model. In the final iteration of our model, we were only able to achieve a 23.01\% BLEU score, which means that our model has low-quality output. This is caused by several factors, from the difficulty to preprocess Tibetan sentences to the lack of Tibetan-English data freely available online and our overall knowledge of the Tibetan language. While it may seem that we were unable to fully accomplish our lofty goal of creating a Tibetan-English translator it’s important to note that we are treading in nearly uncharted territory. We hope that our endeavor may help future projects in this field and provide insight into the current state of Tibetan-English machine translation.

\subsection{Related Work}

@@

To train our model, we acquired publicly available English to Tibetan parallel corpora. The non-profit 84000 provided 80 English to Tibetan parallel texts on their website \cite{c8}. All 80 of these texts were expert translations of Indo-Tibetan Buddhist texts originally written in Tibetan. For our model we used all 80 of these provided texts.

Before using 84000's corpora however, we pre-processed the provided data to meet our model's need. The provided files from 84000 were in a .tmx file format, a very standard parallel corpus file format \cite{c9}. 
For our approach, we scraped all 80 .tmx files for both its English and Tibetan contents and placed them line by line into two large .txt files. One .txt file contained the English lines, and the other .txt file contained the Tibetan translation. We made careful assurance that the lines remained in exactly one to one correspondence with their translations. That is, a newline character was used to distinguish two separate lines of translation, but the same line number in each file corresponds to the exact translation provided by 84000. Therefore, the total line numbers in the two .txt files were necessarily the same since one line in one file corresponding to exactly the same line of translated text in the other file. Two such .txt file were made for each of the 80 .tmx files, a Tibetan and an English .txt file for each. After all 160 .txt files were generated, two single large .txt file were generated by concatenating all the Tibetan .txt files and all the English .txt files. These last two .txt files are what was fed into the model.

@@

\subsection{Model Design}
We used a 6 layer Transformer model that used a multi headed attention mechanism and a switchout transform. Each layer of the transformer model can be divided into an encoder and decoder layer. The encoding layer consists first of a self attention layer, an attention aspect which focuses on the encoded sentence, and then outputs to a feed forward layer. The multi part of the multi headed attention means that each head will encode the information in parallel and then the results will be concatenated and passed to the decoder. Figure one demonstrates this process.
\includegraphics[scale=.4]{Transformer.png}
\caption{\\Figure 1: Overview of a Transformer model \cite{c11}}
\\
\par
To prevent overfitting and to augment the data, we added in a switchout transform which will randomly switch out words in the source and target sentence. This also has the added advantage of augmenting the data even further which is especially important for our problem as religious Tibetan has a relatively small corpus, around 30 thousand words. We also attempted to use masking, however we found it to be ineffective. Masking hides tokens from the model by replacing the tokens with a <mask> token. Tibetan tokens are not separated in the same way that Latin languages separate their tokens. For example,
a line in our test corpus translated to “Having the altruistic intention of excellent bodhicitta.” However, the Tibetan word is only one token. Therefore, the mask would not help the model learn at all, as the entire sentence would be marked as <mask> and there would not be enough context for our model to discover the meaning. As a result, we did not use any form of masking in our model. Switchout does not create the same problem as their is always a token to take its place. \par
After implementing our transformer model with switchout we tuned hyper parameters to maximize our Transformer model. We manually tuned the following hyperparameters: dropout, label smoothing, rnn size, the number of transformer layers and the number of heads of the attention model.  \par
We used four different metrics to evaluate our results: accuracy, predicted score, perplexity and bilingual evaluation understudy (BLEU). The validation accuracy tells us how well we are learning our training data, the predicted score tells us the average log likelihood for each word, the perplexity tells us how well the model predicts the data and the bleu score evaluates the quality of the translation. Together, these metrics provide us with a decent picture of how well our model is performed.

@@

\par
Our six layer transformer model with switch out was unsuccessful at translating religious Tibetan to English. The translated output does not match closely enough in either pure accuracy or in the meaning of the passage to be useful in translating religious Tibetan texts to English. \par
While superficially, the perplexity appears to demonstrate a strong model, further investigation demonstrates that the low perplexity is a sign of inaccuracy. Perplexity measures the number of possible words that a model could predict for any given sentence. A higher perplexity indicates higher uncertainty. However, the low BLEU score, which measures the quality of the machine translation, has already demonstrated that the accuracy is low. This combined with the perplexity shows the models overconfidence in predicting certain phrases. For example, one Tibetan phrases expert translation was “So the bodhisattva mahāsattva Maitreya asked the Blessed One”, but the top five of the predicted translations all are of the form “The Blessed One then said to the [adj] [proper noun]”. Even though this sentence is wrong, the perplexity is low because the model is confident in the first part of the sentence. This phenomenon is repeated throughout the results. Keyphrases from the Tibetan sentence are interpreted wrongly in the same way, and then the model learns to be confident about its predictions which lowers the perplexity. \par
The results demonstrate the flaw in the Transformer model using the current parallel Tibetan to English corpuses. Each token in the Tibetan sentence corresponds to an entire phrase in English rather than a separate word. This complicates the translation, and the model learns parts of the phrases quite well. The validation accuracy is relatively high because the transformer model is good at learning parts of these phrases but it is not good at learning entire sentences or stringing together sentences in context which contributes to the low BLEU score. Overall, the model is capable of learning specific phrases well, but fails at the broad task of creating coherent translations from religious Tibetan texts to English. \par
We are scheduled to discuss these results with our native Tibetan speaker, Palden Gyal, but unfortunately did get to meet in time for this publication.

@@

Considering our use of word segmentation for preprocessing our data (as opposed to character segmentation), our results are comparable with, albeit lower than, Tibetan-Chinese models. Lai, Zhao, and Bao \cite{c2} ran several different Tibetan-Chinese models on the same data using two different preprocessing methodologies, character segmentation and word segmentation. Their word segmented Transformer model averaged a higher BLEU score than ours, scoring 38.44\% while our model scored 23.01\%. While this comparison may lead to the conclusion that our model is flawed and irredeemable, only the former is correct. There’s no doubt that our model is underperforming in comparison to other transformer models and hardly makes a correct prediction at all. However, this is most likely due to our difference in preprocessing techniques and the linguistic differences and similarities between English, Chinese, and Tibetan.

Unlike English but similar to Chinese, Tibetan words are not clearly separated by spaces, rather, characters that represent words are strung together to make a sentence \cite{c5}. This makes preprocessing the data and mapping Tibetan words to their English counterparts a difficult task and it makes preprocessing on characters impossible. We trained our model on a parallel corpus of English and Tibetan sentences instead of segmenting each word since preprocessing these sentences to word-to-word mappings would require extensive knowledge of the Tibetan language which our team did not have. On top of this, a Chinese-Tibetan model might perform overall better than a Tibetan-English one since Chinese and Tibetan are part of the same language family, and it is likely that characters may map to one another and the same preprocessing techniques can be used in both languages. Therefore, while comparing our model to Lai, Zhao, and Bao’s \cite{c2} model may be helpful to give a general baseline on how other Tibetan translators are performing, it should not be considered where our model must perform.

Another issue with our data that may be remedied in future iterations is the corpora we chose to train and test our model on. Since we trained on sentences in our testing set, it is very possible that the model may have read some of the initial characters and assumed that the entire sentence would be a sentence from our training data that began with the same characters. This may explain why our model overpredicts the sentence “Our blessed one” with such a low perplexity. As stated above, this issue might be remedied if we could preprocess our data so that words mapped to words but since we are limited by our ability to do such, we could consider training our model on a different corpus altogether. Instead of using the translated Buddhist sutras to train our model, we could use a more general Tibetan-English corpus to train our model and then use transfer learning to fine-tune our model to religious texts. However, at this current time, the translated sutras provided by 84000 \cite{c8} are some of the only parallel translations available on the internet.

Perhaps the best way to translate Tibetan to English would be to take another approach to our model design altogether and abandon the tested Transformer model. We could build an Encoder-Decoder structure based on Back Translation which would help deal with the lack of Tibetan-English translations. In this model, we would mix Tibetan and English sentences together to train the model \cite{c10}. This would help to increase the size of our training data since it would no longer just be Tibetan sentences. Since this type of model has yet to be tested for Tibetan-English, it's uncertain whether this change could actually improve our translation accuracy and BLEU score, however based on papers written on Back Translation, this design looks promising.

@@

\begin{thebibliography}{99}
